The Companion Mass Distribution to the ELM WD Sample

### 2014-10-08 documents/letter/ms.tex

While the orbital periods and radial velocities indicated the companions were most likely WDs, it was realized that these LMWDs could have NS companions \citep{vLeeuwen07}. However this and other radio searches for pulsed emission, as well as searches for blackbody X-ray emission using {\it XMM-Newton} and {\it Chandra}, have all been fruitless \citet{agueros09a,agueros09b,kilic13}. Yet, such systems must exist since LMWDs are observed as companions around millisecond pulsars. Since the orbital solutions provided by the ELM survey combined with the precision of radio data would overconstrain the system, finding even one system would be extraordinarily interesting. The NS binaries with WD companions bright enough to detect spectroscopic radial velocity variations have precisely determined NS masses, allowing for an understanding of the binary evolution processes forming these systems \citep{vKerkwijk96,callanan98,bassa06,antoniadis12}. The resources required for these targeted observations, selecting the most likely candidates for multiwavelength follow-up is crucial. We outline the standard technique below.

@@

\begin{equation}
P({\rm NS}) = \int_{0^{\circ}}^{i_{\rm NS}} {\rm sin}\ i\ {\rm d}i
\end{equation}

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[angle=90,width=1\columnwidth]{model_post.eps}
\caption{Replace this text with your caption}
\end{center}
\end{figure}

@@

We apply our model to the ELM WD sample, however there are some slight differences worth mentioning in the population of ELM WDs and our test cases above.

First, there are 55 systems with radial velocity measurements in the ELM sample, while the test cases above were each produced with 100 systems. Second, the masses of the primaries in our test cases were all $0.25~\Msun$, while in the ELM sample, the masses vary between 0.16 and $0.49~\Msun$. Also, the orbital periods of the systems in our test case were all 12 hours, while in the ELM sample the orbital periods vary between 12 minutes and just over a day. Since our model does not depend on the $M_2$ or $P_{\rm orb}$, its accuracy will not be affected.

In the ELM sample three systems are eclipsing binaries, with known companion masses (J0345$+$1748: $M_2=0.72~\Msun$, J0651$+$2844: $M_2=0.50~\Msun$, and J0751$-$0141: $M_2=0.97~\Msun$) {\bf Cite something here}. These are taken into account in our model through equation {\bf XXX}. Since the inclination angle cannot vary, these systems will strongly affect the likelihood of any particular model.

Finally, six systems in the ELM WD sample lack radial velocity variations down upper limits of $\sim$50 km s$^{-1}$. While some of these may be in binaries with radial velocities below the detection limit, they may also be single ELM WDs. In principle we could add to our model the possibility that each of these WDs is actually a single ELM WD. Because this requires an additional six free parameters, one for each ELM WD without radial velocity variations, including these in our model becomes too computationally expensive. The result is that we can run our model on the 55 ELM systems with orbital solutions.

We apply our model to the ELM WD data set, using 10 independent Markov chains, each run for 100,000 steps. The acceptance fraction is {\bf XXX}. We determine an autocorrelation length of {\bf XXX} steps, and correspondingly, only include every {\bf XXX} samples for our posterior distributions. {\bf Is there anything else we need to say about the applying the method to the data set?}

The results of our model are shown in Figure {\bf XXX}. The first panel shows the distribution of $M_{2,min}$ for the data set. It is fairly similar to the distribution of $M_{2,min}$ in the test cases above, with a small high mass component, and the majority of sample with $M_{2,min}$ below $0.8~\Msun$. The second panel shows the posterior distribution of $\mu$ and $\sigma$; contours indicate 1, 2, and 3-$\sigma$ confidence levels. The single high probability region indicates the models converged {\bf (Is this reasonable to say?)}. The best fit Gaussian model has a mean of 0.66 $\substack{+0.12 \\ -0.09}~\Msun$ and a standard deviation of 0.27$\substack{+0.09 \\ -0.07}~\Msun$. The third panel shows the NS fraction is strongly peaked toward low probabilities. Our model indicates a NS fraction of 1.4$\substack{+7.0 \\ -1.4}$\% at the 68\% confidence level, however there is a non-negligible tail toward higher NS probabilities. This is further indicated by the fourth panel in Figure {\bf XXX} which shows a histogram of NS probabilities for individual systems. These probabilities are obtained by taking the average of the posterior distribution of NS probabilities for each system. Although a few systems have high probabilities, the distribution is strongly clustered below 20\%.

@@

How does the companion mass distribution compare to the single WD mass distribution? Based on their population of $\sim$ 14,000 spectroscopically confirmed DA WDs in SDSS, \citet{kleinman13} found that WDs are well approximated by a four component Gaussian mixture model. Their mixture model is dominated by two Gaussians with means $\sim0.6~\Msun$, containing over 75\% of all WDs. Figure {\bf XXX} shows that the best fit Gaussian to our distribution of WD companion masses is also centered around $\sim0.6!\Msun$. However, our distribution is significantly wider ($\sigma \approx 0.3~\Msun$ compared with $\sigma \approx 0.1~\Msun$). This could be due to our simplistic assumption of a single Gaussian mass distribution. A more complicated model would include multiple Gaussians with varying weights, however adding only one more Gaussian component doubles the number of free parameters from three to six. Finding the maximum likelihood in a three dimensional parameter space is computationally expensive, and moving to a six dimensional space may require more complex numerical routines.

There is no reason to expect the true companion mass distribution to ELM WDs follows a Gaussian curve. However, the posterior distribution in $\mu$ and $\sigma$ is well behaved, with a clear high probability region, similar to the posterior distributions for $\mu$ and $\sigma$ for our first two test cases shown in Figure {\bf XXX}. The part of the distribution with a low $\mu$ and high $\sigma$ is indicative of a low mass WD component. A low mass component is expected from binary evolution theory; recent population synthesis work predicts more than half of all He WDs in DWDs will have He WD companions \citet{toonen12}. Our results here suggest that this is not the case, our posterior suggests a Gaussian distribution centered around $0.6~\Msun$, firmly in the mass range for CO WDs. This suggests that our results on the companion masses of the secondaries in these systems could constrain population synthesis models. We plan on exploring this in a future work.

Systems with NS companions cannot be differentiated from those with WD companions through orbital period and ELM WD mass alone. Our model attempts to suggest a NS fraction for the population as a whole. This is an improvement over NS probabilities for individual companions calculated through equation {\bf XXX}. The third panel in Figure {\bf XXX} indicates a NS fraction of a few percent. For 55 systems with observed radial velocity variations, this corresponds to a 1-2 systems with NS companions. However, the top left panel in Figure {\bf XXX} shows that even a system with $M_{2,min}=1.0~\Msun$ can be a WD from the high mass tail of a Gaussian distribution. Furthermore, in their Gaussian mixture fit to the SDSS DA WD mass distribution, \citet{kleinman13} found a high mass WD component extending above $1~\Msun$. Our single Gaussian model does not include this high mass component. It is not guaranteed that any ELM WDs have NS companions. Nevertheless, a quite firm result of our model is an upper limit on the fraction of ELM WDs with NS companions of {\bf XXXX}\%.

@@

The fourth panel in Figure {\bf XXX} shows  that only two systems have $P(NS)$ above 50\%: J0811$0225 (67\%) and J1741$6526 (53\%). These systems are ideal candidates for searches for NS companions in radio and X-ray. Radio observations of these objects are in progress, however because a NS companion may be radio quiet or beamed away from us, even if this system is a NS, it may not be detectable in radio. However, all NS give off blackbody radiation as X-rays, and therefore X-ray follow-up provides a more stringent test on the potential for a NS companion.

Non-detections? What does that imply about the existence of single ELM WDs?

\begin{figure}[h!]
\begin{center}
\includegraphics[angle=90,width=1\columnwidth]{ELM_post.eps}
\caption{Replace this text with your caption}
\end{center}
\end{figure}

>>>>

While the orbital periods and radial velocities indicated the companions were most likely WDs, it was realized that these LMWDs could have NS companions \citep{vLeeuwen07}. However this and other radio searches for pulsed emission, as well as searches for blackbody X-ray emission using {\it XMM-Newton} and {\it Chandra}, have all been fruitless \citep{agueros09a,agueros09b,kilic13}. Yet, such systems must exist since LMWDs are observed as companions around millisecond pulsars. Since the orbital solutions provided by the ELM survey combined with the precision of radio data would overconstrain the system, finding even one system would be extraordinarily interesting. The NS binaries with WD companions bright enough to detect spectroscopic radial velocity variations have precisely determined NS masses, allowing for an understanding of the binary evolution processes forming these systems \citep{vKerkwijk96,callanan98,bassa06,antoniadis12}. The resources required for these targeted observations, selecting the most likely candidates for multiwavelength follow-up is crucial. We outline the standard technique below.

@@

\begin{equation}
P({\rm NS}) = \int_{0^{\circ}}^{i_{\rm NS}} {\rm sin}\ i\ {\rm d}i \label{eq:P_NS_approx}
\end{equation}

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[angle=90,width=1\columnwidth]{model_post.eps}
\caption{Replace this text with your caption}
\label{fig:tests}
\end{center}
\end{figure}

@@

The ELM WD sample is composed of 55 systems with radial velocity variations fit to orbital solutions, providing precise measurements for $P_{\rm orb}$ and $K$. Fits to spectroscopic templates provide high S/N spectra provide precisely determined $M_1$. We use the spectroscopic solutions found in \citet{Gianninas14}. Furthermore, three systems are eclipsing binaries, with known companion masses (J0345$+$1748: $M_2=0.72~\Msun$, J0651$+$2844: $M_2=0.50~\Msun$, and J0751$-$0141: $M_2=0.97~\Msun$). These are taken into account in our model through equation {\bf XXX}. Since the inclination angle cannot vary, these systems will strongly affect the likelihood of any particular model.

Finally, six systems in the ELM WD sample lack radial velocity variations, with upper limits of $\approx$20-50 km s$^{-1}$. Some of these may be in binaries with radial velocities below the detection limit, due to low inclination angle, however studies by {\bf XXX} suggest there may also be a population of single LMWDs. Because of the uncertainty in their formation, we restrict these from our sample.

%In principle we could add to our model the possibility that each of these WDs is actually a single ELM WD. Because this requires an additional six free parameters, one for each ELM WD without radial velocity variations, including these in our model becomes too computationally expensive.

%We apply our model to the ELM WD data set, using 10 independent Markov chains, each run for 100,000 steps. The acceptance fraction is {\bf XXX}. We determine an autocorrelation length of {\bf XXX} steps, and correspondingly, only include every {\bf XXX} samples for our posterior distributions. {\bf Is there anything else we need to say about the applying the method to the data set?}

The results of our model are shown in Figure \ref{fig:ELM_post}. The first panel shows the distribution of $M_{2,min}$ for the data set, showing a small high mass component, and the majority of sample with $M_{2,min}$ below $0.8~\Msun$. The second panel shows the posterior distribution of $\mu$ and $\sigma$; contours indicate 1, 2, and 3-$\sigma$ confidence levels. The best fit Gaussian model has a mean of 0.66 $\substack{+0.12 \\ -0.09}~\Msun$ and a standard deviation of 0.27$\substack{+0.09 \\ -0.07}~\Msun$. The third panel shows the NS fraction is strongly peaked toward low probabilities. Our model indicates a NS fraction of 1.4$\substack{+7.0 \\ -1.4}$\% at the 68\% confidence level, however there is a non-negligible tail toward higher NS probabilities. This is further indicated by the fourth panel in Figure \ref{fig:ELM_post} which shows a histogram of NS probabilities for individual systems. These probabilities are obtained by taking the average of the posterior distribution of NS probabilities for each system. Although a few systems have high probabilities, the distribution is strongly clustered below 20\%.

@@

Our results indicate that the best fit Gaussian for the companions to the ELM WD sample is very similar to that of the population of single DA WDs in SDSS with a mean of 0.6 $\Msun$ \citep{kleinman13}. However, our distribution is significantly wider ($\sigma \approx 0.3~\Msun$ compared with $\sigma \approx 0.1~\Msun$), likely due to mass transfer affecting the mass distribution of the unseen primary WDs as well. Furthermore, population synthesis predictions suggest there may be two distinct companion populations to LMWD, composed of both CO and He core WDs. The ratio of these two populations depends on the exact stellar evolution prescriptions assumed \citep{toonen12}. Our method here could be extended to include multiple Gaussians with varying weights, however adding only one more Gaussian component doubles the number of free parameters from three to six. Finding the maximum likelihood in a three dimensional parameter space is computationally expensive, and moving to a six dimensional space may require more complex numerical routines. In principle, the ratio of He+He to He+CO WD systems could be determined using two Gaussian model, potentially placing an important constraint on population synthesis models. It remains to be seen whether such a model can place such constraints with 55 measured systems. We plan to explore this in a future work, however it may require a larger data set.

There is no reason to expect the true companion mass distribution to ELM WDs follows a Gaussian curve. However, the posterior distribution in $\mu$ and $\sigma$ is well behaved, with a clear high probability region, similar to the posterior distributions for $\mu$ and $\sigma$ for our first two test cases shown in Figure \ref{fig:ELM_post}. The part of the distribution with a low $\mu$ and high $\sigma$ is indicative of a low mass WD component.

Systems with NS companions cannot be differentiated from those with WD companions through orbital period and ELM WD mass alone. Our model attempts to determine a NS fraction for the population as a whole. This is an improvement over NS probabilities for individual companions calculated through Equation \ref{P_NS_approx}. The third panel in Figure \ref{fig:ELM_post} indicates a NS fraction of a few percent. For 55 systems with observed radial velocity variations, this corresponds to a 1-2 systems with NS companions. However, the top left panel in Figure \ref{fig:tests} shows that even a system with $M_{2,min}=1.0~\Msun$ can be a WD from the high mass wing of our Gaussian distribution. It is not guaranteed that any of the LMWDs in the ELM WD sample have NS companions. Nevertheless, our model indicates a NS fraction of {\bf XXX\%}, corresponding to zero to four systems.

@@

The fourth panel in Figure {\bf XXX} shows  that only two systems have $P(NS)$ above 25\%: J0811$+$0225 ({\bf XXX}\%) and J1741$+$6526 ({\bf XXX}\%). These systems are ideal candidates for searches for NS companions in radio and X-ray. Radio observations of these objects are in progress, however because a NS companion may be radio quiet or beamed away from us, even if this system is a NS, it may not be detectable in radio. However, all NS give off blackbody radiation as X-rays, and X-ray follow-up provides a more stringent test on the potential for a NS companion.

\begin{figure}[h!]
\begin{center}
\includegraphics[angle=90,width=1\columnwidth]{ELM_post.eps}
\caption{Replace this text with your caption}
\label{fig:ELM_post}
\end{center}
\end{figure}
